Lattice Boltzmann study of mass transfer for two-dimensional Bretherton/Taylor bubble train flow

### 2012-12-24 Text/Final/introduction.tex
After revision - submission is done. Hopefully, it's final.

Despite their simplicity, such analytical expressions work well for flows with low
capillary numbers $Ca<0.1$ \cite{bercic-mass} where the bubble shape is
symmetrical and can be approximated with good precision. Moreover, because of the
hydrodynamic pattern in the slug (i.e. presence of a vortex in slug), one can
estimate the time for a fluid batch to travel the whole circumference.   However,
with the increase of the capillary number the situation changes significantly -- the
symmetrical bubble shape is lost and the bubble resembles a bullet \cite{kuzmin-binary2d}.
For flows with $Ca>0.7$ there is also no vortex in the liquid slug. In this case the Higbie theory fails
to estimate the contribution from bubble caps, which explains the need to turn to numerical simulations where all
hydrodynamic fields as well as complex bubble shapes are taken into account.

Typical numerical studies of mass transfer
\cite{kreutzer-overview,vanbaten-circular} do not consider the simulation of
bubble shapes for $Ca>0.1$. {\color{red} The interesting work by
\citet{onea-mass} only takes into account short length bubbles in microchannels
with a square cross section. As well, in this work  the Schmidt number around $0.8$ is  used:
thus, the thickness of the concentration and momentum boundary layer is similar
leading to qualitative results.} The usual simulation of mass transfer is
performed as follows: 
\begin{description}
 \item[I] The bubble shape is calculated either
through analytical correlations \cite{bretherton} or experimental correlations
\cite{cerro-bubble-train} without directly resolving bubble shapes through
multiphase simulations.  The expressions for bubble shapes are available only
for flows with capillary number $Ca<0.1$.
  \item[II] Hydrodynamic fields
are then obtained by performing simulations of one-component flow around the
bubble by imposing the bubble velocity on the channel walls. Thus, the simulations are
performed in the reference frame moving with the bubble. A stress-free
condition is imposed at the bubble surface.
  \item[III]  The mass transfer
simulations are performed in the reference frame moving with the bubble. The
saturation concentration is imposed at the bubble surface. Only one unit cell
containing a single bubble is used for simulations. Periodic concentration boundary conditions are
utilized to determine the volumetric mass transfer coefficient, which is
calculated through the following equation \cite{vanbaten-circular}:
\begin{equation} \label{main:simulation:equation}
\vol=\frac{\mathrm{\overline{Flux}}}{\cbubble-\langle\cinoutlet\rangle}
\frac{\mathrm{bubble\ surface\ area}}{\mathrm{unit\ cell\ volume}},
\end{equation}
 where $\langle\cinoutlet(t)\rangle=\int{C \uinoutlet
\mathrm{d}A}/\int{\uinoutlet\mathrm{d}A}$ is the space-averaged inlet/outlet (periodic boundary conditions)
concentration as a function of time. Therefore, in terms of the mass transfer definition, $\langle\cinoutlet(t)\rangle$ plays the role of the 
characteristic concentration.
%This is the characteristic concentration in the definition, Eq. \ref{eq:main:definition}.  
The time-averaged
concentration flux ($\mathrm{\overline{Flux}}$) is calculated as the difference
between the overall average concentration in the whole domain
($\langle\coverall\rangle=\int_{V} C \mathrm{d}V /V$) at time $t_1$ and at time
$t_2$ divided by the time difference $t_2-t_1$. The agreement between numerical
simulations \cite{vanbaten-circular} and experimental correlations of \citet{bercic-mass}
% TODO: was good where?
was good. 
 \end{description}

@@

Addressing situations for a rich number of hydrodynamic patterns, shapes, and
effects of bubble lengths, etc for bubble train flows, we feel that there is a need to examine 
carefully the strategies and assumptions behind the numerical calculations
of the mass transfer coefficient. We aim at establishing clear
procedures as to how properly obtain the mass transfer coefficient via a study
of different boundary conditions and different definitions of the characteristic concentration.
The case we want to examine is a two-dimensional bubble train flow between parallel plates.
We address the following issues:
 \begin{description}
 \item[I] Applicability of periodic boundary
conditions to determine the mass transfer coefficient when the vortex in the slug
disappears, i.e. when $Ca>0.7$.
 \item[II] Validity of the inlet/outlet-averaged or domain-averaged concentrations as
characteristic concentrations in the definition of the mass transfer
coefficient.
 \item[III] Translation of the continuous experimental picture to
numerical simulations of a few unit cells, the issue of correspondence
between space averages (simulations \cite{vanbaten-circular}) and time averages (experiment).
% \item[IV] Influence of the hydrodynamic pattern in flows with $Ca>0.7$ on the mass transfer.
\end{description}
In addition, at the end of the manuscript we present results of the dependence of the volumetric
mass transfer coefficient on the Peclet number that we compare with analytical \cite{irandoust} and experimental correlations
\cite{yue-mass}.  The thorough determination of the mass transfer coefficient and associated Sherwood number as a
function of other non-dimensional parameters such as gas holdup, bubble/slug
lenghts, and the capillary number is left for future studies.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{benchmark_hydro.eps}
\caption{Simplified sketch of the bubble motion. Using periodic
conditions for the velocity field is natural, but needs evaluation for mass
transfer.
\label{fig:benchmark:hydro}}
\end{figure}

@@

The paper is organized as follows. We start with definitions of the volumetric mass transfer coefficient and apply them 
to the bubble train flow to derive expressions to connect the space- and time-averages.  Then, the lattice
Boltzmann model used to simulate mass transfer is presented, followed by benchmarks. Finally,
numerical simulations of various boundary conditions and simulations spanning a few unit cells for
different hydrodynamic patterns are presented to establish the procedure to determine the
volumetric mass transfer coefficient. The comparison with analytical correlations is also discussed.

\section{Mass transfer definitions}
\label{sec:mass:definitions}
By definition, the mass transfer coefficient from a surface with an imposed constant
concentration $\cbubble$ is:
\beqal
\label{eq:main:definition}
&k_L=\frac{\dot{m}}{P \Delta C},\ \Delta C= \cbubble - \cmedium,
\feqal
where $\dot{m}$ is the mass flux $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, and $\Delta C$ is the concentration difference between the surface and the surrounding medium
$\Bigl[\frac{kg}{m^3}\Bigr]$. Therefore, $k_L$ has a dimension of 
$\Bigl[\frac{m}{s}\Bigr]$. Usually, the surrounding medium concentration is taken at an infinite distance
from the bubble. However, in the case of complicated geometries and non-homogeneous concentrations, 
the medium concentration can be the average concentration in the domain or the flux-averaged
concentration at the inlet or outlet, etc. Thus, one needs to establish a clear definition of  $\Delta C$ to determine the volumetric
mass transfer coefficient in the case of complex geometries and non-trivial hydrodynamic velocity patterns.

@@

If one can assume that the point mass sources are distributed in the whole
medium, the mass accumulated in the volume $V=A \Delta x$ can be
calculated as the difference of mass fluxes entering and leaving the domain $U
\bigl(C(x+\Delta x)-C(x)\bigr)$. The accumulated mass should be proportional to
the mass transfer coefficient:
\beq
U \bigl(C(x+\Delta x)-C(x)\bigr)=k_L P \bigl(\cstar-C(x)\bigr), 
\feq 
giving the same solution but only in the spatial domain:
\beq
%\label{concentration:domain:coordinate}
C(x)= \cstar \Bigl(1-\exp\bigl(-k_L a \frac{x}{U} \bigr)\Bigr).
\label{main:mass:transfer:expression} 
\feq
Note that the concentration $C(x)$ does not depend on time.

\item If one transfers to the frame moving with the liquid velocity $U$, the situation will be
the same as in the first case. One can connect  time and space with the
velocity $U$ ($t=\frac{x}{U}$) to obtain the same equation as in the case 2.
\end{enumerate}

\subsection{Bubble train}
In the application to the bubble train flow it is useful to think of one bubble as a point
source to be able to use the calculations presented above. For example, the expression
(\ref{main:mass:transfer:expression}) was used in experiments by
\citet{bercic-mass}. However, one should be accurate with the definition of velocities because two
different phases co-exist in the bubble train flow. Usually, one can take the velocity $U$ to be
a bulk velocity or $U=\ugas+\uliq$, where $\ugas$ and $\uliq$ are liquid and gas
superficial velocities, respectively. {\color{red} The gas and liquid superficial and actual velocities are connected with each other through a complex dependency on $Ca$, $\epsilon$ \cite{kuzmin-binary3d}.}

With experimental measurements of concentration at different locations, the calculation
of the mass transfer coefficient using the logarithmic function is straightforward.
However, if one wants to analytically or numerically calculate the mass transfer coefficients, the
situation is much more complicated because of the presence of two phases and complex bubble
geometry. As was mentioned before, depending on the capillary
number the velocity pattern and thus scalar mixing is different. Analytical approaches
\cite{irandoust,vanbaten-circular} assume that the
contributions from film and bubble caps can be calculated separately. Therefore no tracer from the film influences bubble caps diffusion.   However, this assumption overpredicts mass transfer for a number of experiments \cite{irandoust}. This happens since some tracer
concentration from the film is mixed with the slug and increases the overall concentration in the slug, thereby decreasing
the mass transfer from the bubble caps.
Therefore, the analytical estimates for the mass transfer coefficient calculation  do not account for mutual mass
transfer from neighbouring bubbles.

@@

We perform simulations in the frame co-moving with the bubble
in which the bubble position stays constant. The bubble velocity $\ububble$ is
different from the bulk velocity $U=\ugas+\uliq$, and one thus needs to perform a $x$ coordinate
variable change:
\beqal
\label{theor:average:concentration:time}
&x(t)=\ububble t\\
&\overline{C(x)}=\cstar \Bigl(1-\exp\bigl(-\vol \frac{x}{\ugas+\uliq}\bigr)\Bigr)\\
&\langle C(t)\rangle=\cstar \Bigl(1-\exp\bigl(-\vol\, t \frac{\ububble}{\ugas+\uliq}\bigr)\Bigr),
\feqal
where $\langle C(t)\rangle$ is the space-averaged characteristic concentration,
and $\overline{C(x)}$ is the time-averaged concentration at location $x$. 
%The geometry is not one-dimensional and one needs the characteristic concentration
%to be only as a function of time. 
One can make different choices for
$\langle C(t) \rangle$ such as the concentration averaged over the whole domain or inlet/outlet space-averaged
concentrations used in works \cite{vanbaten-circular,kreutzer-overview}.  
%Note that $C(x)$ is not averaged in time as $\overline{C(x)}$ since $C(x)$
%ideally should not depend on time. 
The volumetric mass transfer coefficient can be obtained through the space-averaged concentration:
\beqal
\label{theor:one:concentration:time}
&\vol\, t \frac{\ububble}{\ugas+\uliq}=\ln \frac{\cstar}{\cstar-\langle C(t)\rangle}\\
&\volnondim=\frac{\lunit}{\ububble t}\ln \frac{\cstar}{\cstar-\langle C(t) \rangle},
\feqal
where the parameter $\vol \frac{\lunit}{\ugas+\uliq}$ is non-dimensional. One can also measure the
volumetric mass transfer coefficient from concentrations given at times $t_1$ and $t_2$:
\beq
\label{theor:continuous:mass:transfer}
\volnondim=\frac{\lunit}{\ububble
(t_2-t_1)}\ln\frac{C^{*}-\langle C(t_1) \rangle}{C^{*}-\langle C(t_2) \rangle}.
\feq
Expressions (\ref{theor:average:concentration:time} - \ref{theor:continuous:mass:transfer}) are the
cornerstones of the present work . Four possible scenarios of numerical simulations have been examined: 
\begin{enumerate}
\item % The bubble train multiphase simulations are conducted by applying periodic boundary
%conditions \cite{kuzmin-binary2d,kuzmin-binary3d}. 
%One
%can make an assumption of same boundary
%conditions applied to the tracer concentration to simulate mass transfer. 
One unit
cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. In this case no tracer leaves the
domain similarly to the plug flow. 
%From the physical point of view this case resembles the mass transfer for small Capillary numbers
%$Ca<0.1$.  The liquid slug is well mixed and one have a homogeneous concentration in the unit
%cell, i.e. the outlet and the inlet concentrations are close. 
Though easier to implement, it gives rise to the criticism
that the inlet concentration is equal to the outlet one. As was
discussed, in experiments there is a concentration difference between the
inlet and the outlet, even for one unit cell.

In this case, the volumetric mass transfer coefficient is calculated by Eq. \ref{theor:one:concentration:time}.
The
characteristic concentration $\langle C(t) \rangle$ required for the volumetric mass
transfer coefficient is taken as the average concentration in the domain:
\beq
\label{average:concentration:definition}
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\feq
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{benchmark_middle.eps}\\
\caption{The two-dimensional benchmarks for the  the mass transfer
coefficient (bottom) for the bubble located near the entrance (top) and at the middle of the domain
(bottom). \label{fig:benchmark}}
\end{figure}
\item Periodic boundary conditions are applied as in the first case but the characteristic concentration is taken as the inlet/outlet flux-averaged
concentration \cite{vanbaten-circular}:
\beqal
&\langle \cinlet(t) \rangle=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
&\langle \coutlet(t) \rangle=\frac{\int{U(y) C(\lunit,y,t) \mathrm{d}y}}{\int{U(\lunit,y)
\mathrm{d}y}}\\
&\cinlet(\bm{x},t)=\coutlet(\bm{x},t),\,\text{due to periodicity}.
\feqal
The assumptions of this approach are
that the concentration difference between the inlet/outlet- and the space-averaged over the whole unit cell is not significant. Thus, the tracer is assumed to be well mixed in the slug.
\item
The approach of \citet{vanbaten-circular}, where periodic boundary conditions are used and
the mass transfer coefficient is calculated as the gain of the mass in the system
divided by the concentration difference
multiplied by the surface area:
\beq
\label{eq:vanbaten:formulation}
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-\langle C(t) \rangle)},
\feq
where the mass flux in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(\bm{x},t_2)\mathrm{d}\bm{x}}-\int_{liq}{
C(\bm{x},t_1)\mathrm { d } \bm{x}} } {
t_2-t_1 }.
\feq
In the approach of \citeauthor{vanbaten-circular} the inlet/outlet flux-averaged concentrations were taken as the
characteristic concentration $\langle C(t) \rangle $.

@@

Periodic boundary conditions are the extreme
limiting case of Eq. \ref{main:main:main}: 
\beqstar
\int{\coutlet(t)
u(\lunit,y)\mathrm{d}y}=\int{\cinlet(t) u(0,y)\mathrm{d}y}. 
\feqstar
%Then the mass change in the domain arises only because of the bubble
%surface. However, to calculate the mass transfer coefficient one needs to have
%the characteristic concentration  $\langle C(t) \rangle$ and the mass gain $\dot{m}$
%which do require the storage of both temporal and spatial information.
Another limiting case (will be shown later) is when the mass accumulation rate equals zero, i.e. $\dot{m}=0$. 
This situation corresponds to a simulation of a few unit cells with open boundary for flows with
$Ca>0.7$.   
%One
%can use only spatial information to calculate the mass transfer coefficient,
%i.e. $\int{\cinlet(t) u(0,y) \mathrm{d}y}$ and $\int{\coutlet(t)
%u(\lunit,y)\mathrm{d}y}$ require only space information.

@@

\subsection{TRT $D2Q9$ model}
The lattice Boltzmann equation (LBE) operates on a square/cubic grid representing the
physical domain {\color{red} in the non-dimensional space} (hereafter we use subscript LB to denote non-dimensional quantities). It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. The LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along  
directions specified by a discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)= f_i(\bm{x},t)-\omega \bigl(f_i(\bm{x},t)-eq_i(\bm{x},t)\bigr),&&\text{
collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $eq_i$ is the equilibrium probability distribution function, and $\omega$ is the
relaxation parameter. The term $-\omega \bigl(f_i- eq_i\bigr)$ is the so-called BGK
collision operator \cite{bgk}. However, the approach used here is the TRT
(two-relaxation-times) collision operator \cite{ginzburg-main,ginzburg-saturated-flow}. In
comparison with the widely used BGK collision operator, the TRT collision operator has better accuracy
for diffusion and convection fluxes, as well as a larger range of parameters where the scheme is stable.

>>>>

Despite their simplicity, such analytical expressions work well for flows with low
capillary numbers $Ca<0.1$ \cite{bercic-mass} where the bubble shape is
symmetrical and can be approximated with good precision. Moreover, because of the
hydrodynamic pattern in the slug (i.e. presence of a vortex in the slug), one can
estimate the time for a fluid batch to travel the whole circumference.   However,
with the increase of the capillary number the situation changes significantly -- the
symmetrical bubble shape is lost and the bubble resembles a bullet \cite{kuzmin-binary2d}.
For flows with $Ca>0.7$ there is also no vortex in the liquid slug. In this case the Higbie theory fails
to estimate the contribution from bubble caps, which explains the need to turn to numerical simulations where all
hydrodynamic fields as well as complex bubble shapes are taken into account.

Typical numerical studies of mass transfer
\cite{kreutzer-overview,vanbaten-circular} do not consider the simulation of
bubble shapes for $Ca>0.1$. {\color{red} The interesting work by
\citet{onea-mass} only takes into account short length bubbles in microchannels
with a square cross section. As well, in this work  a Schmidt number of around $0.8$ is  used:
thus, the thickness of the concentration and momentum boundary layer is similar
leading to qualitative results since it's not possible to separate mixing due to hydrodynamics patterns from mixing due to diffusion.}

The usual simulation of mass transfer is
performed as follows: 
\begin{description}
 \item[I] The bubble shape is calculated either
through analytical correlations \cite{bretherton} or experimental correlations
\cite{cerro-bubble-train} without directly resolving bubble shapes through
multiphase simulations.  The expressions for bubble shapes are available only
for flows with capillary number $Ca<0.1$.
  \item[II] Hydrodynamic fields
are then obtained by performing simulations of one-component flow around the
bubble by imposing the bubble velocity on the channel walls. Thus, the simulations are
performed in the reference frame moving with the bubble. A stress-free
condition is imposed at the bubble surface.
  \item[III]  The mass transfer
simulations are performed in the reference frame moving with the bubble. The
saturation concentration is imposed at the bubble surface. Only one unit cell
containing a single bubble is used for simulations. Periodic concentration boundary conditions are
utilized to determine the volumetric mass transfer coefficient, which is
calculated through the following equation \cite{vanbaten-circular}:
\begin{equation} \label{main:simulation:equation}
\vol=\frac{\mathrm{\overline{Flux}}}{\cbubble-\langle\cinoutlet\rangle}
\frac{\mathrm{bubble\ surface\ area}}{\mathrm{unit\ cell\ volume}},
\end{equation}
 where $\langle\cinoutlet(t)\rangle=\int{C \uinoutlet
\mathrm{d}A}/\int{\uinoutlet\mathrm{d}A}$ is the space-averaged inlet/outlet (periodic boundary conditions)
concentration as a function of time. Therefore, in terms of the mass transfer definition, $\langle\cinoutlet(t)\rangle$ plays the role of the 
characteristic concentration.
%This is the characteristic concentration in the definition, Eq. \ref{eq:main:definition}.  
The time-averaged
concentration flux ($\mathrm{\overline{Flux}}$) is calculated as the difference
between the overall average concentration in the whole domain
($\langle\coverall\rangle=\int_{V} C \mathrm{d}V /V$) at time $t_1$ and at time
$t_2$ divided by the time difference $t_2-t_1$. The agreement between numerical
simulations \cite{vanbaten-circular} and experimental correlations of \citet{bercic-mass}
% TODO: was good where?
was good. 
 \end{description}

@@

Addressing situations for a rich number of hydrodynamic patterns, shapes, and
effects of bubble lengths, etc for bubble train flows, we feel that there is a need to examine 
carefully the strategies and assumptions behind the numerical calculations
of the mass transfer coefficient. We aim at establishing clear
procedures as to how properly obtain the mass transfer coefficient via a study
of different boundary conditions and different definitions of the characteristic concentration.
The case we want to examine is a two-dimensional bubble train flow between parallel plates.
We address the following:
 \begin{description}
 \item[I] Applicability of periodic boundary
conditions to determine the mass transfer coefficient when the vortex in the slug
disappears, i.e. when $Ca>0.7$.
 \item[II] Validity of the inlet/outlet-averaged or domain-averaged concentrations as
characteristic concentrations in the definition of the mass transfer
coefficient.
 \item[III] Translation of the continuous experimental picture to
numerical simulations of a few unit cells, the issue of correspondence
between space averages (simulations \cite{vanbaten-circular}) and time averages (experiment).
% \item[IV] Influence of the hydrodynamic pattern in flows with $Ca>0.7$ on the mass transfer.
\end{description}
In addition, at the end of the manuscript we present results of the dependence of the volumetric
mass transfer coefficient on the Peclet number that we compare with analytical \cite{irandoust} and experimental correlations
\cite{yue-mass}.  The thorough determination of the mass transfer coefficient and associated Sherwood number as a
function of other non-dimensional parameters such as gas holdup, bubble/slug
lenghts, and the capillary number is left for future studies.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{benchmark_hydro.eps}
\caption{Simplified sketch of the bubble motion. Using periodic
conditions for the velocity field is natural, but needs evaluation for mass
transfer.
\label{fig:benchmark:hydro}}
\end{figure}

@@

The paper is organized as follows. We start with definitions of the volumetric mass transfer coefficient and apply them 
to  bubble train flow in order to derive expressions to connect space- and time-averages.  Then, the lattice
Boltzmann model used to simulate mass transfer is presented, followed by benchmarks. Finally,
numerical simulations of various boundary conditions and simulations spanning a few unit cells for
different hydrodynamic patterns are presented to establish the procedure to determine the
volumetric mass transfer coefficient. The comparison with analytical correlations is also discussed.

\section{Mass transfer definitions}
\label{sec:mass:definitions}
By definition, the mass transfer coefficient from a surface with an imposed constant
concentration $\cbubble$ is:
\beqal
\label{eq:main:definition}
&k_L=\frac{\dot{m}}{P \Delta C},\ \Delta C= \cbubble - \cmedium,
\feqal
where $\dot{m}$ is the mass flux $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, and $\Delta C$ is the concentration difference between the surface and the surrounding medium
$\Bigl[\frac{kg}{m^3}\Bigr]$. Therefore, $k_L$ has units of 
$\Bigl[\frac{m}{s}\Bigr]$. Usually, the surrounding medium concentration is taken at an infinite distance
from the bubble. However, in the case of complicated geometries and non-homogeneous concentrations, 
the medium concentration can be the average concentration in the domain or the flux-averaged
concentration at the inlet or outlet, etc. Thus, one needs to establish a clear definition of  $\Delta C$ to determine the volumetric
mass transfer coefficient in the case of complex geometries and non-trivial hydrodynamic velocity patterns.

@@

If one can assume that the point mass sources are distributed in the whole
medium, the mass accumulated in the volume $V=A \Delta x$ can be
calculated as the difference of mass fluxes entering and leaving the domain $U
\bigl(C(x+\Delta x)-C(x)\bigr)$. The accumulated mass should be proportional to
the mass transfer coefficient:
\beq
U \bigl(C(x+\Delta x)-C(x)\bigr)=k_L P \bigl(\cstar-C(x)\bigr), 
\feq 
giving the same solution, now in the spatial domain:
\beq
%\label{concentration:domain:coordinate}
C(x)= \cstar \Bigl(1-\exp\bigl(-k_L a \frac{x}{U} \bigr)\Bigr).
\label{main:mass:transfer:expression} 
\feq
Note that the concentration $C(x)$ does not depend on time.

\item If one transfers to the frame moving with the liquid velocity $U$, the situation will be
the same as in the first case. One can connect  time and space with the
velocity $U$ ($t=\frac{x}{U}$).  %to obtain the same equation as in the case 2.
\end{enumerate}

\subsection{Bubble train}
In the application to bubble train flow it is useful to think of one bubble as a point
source to be able to use the calculations presented above. For example, the expression
(\ref{main:mass:transfer:expression}) was used in experiments by
\citet{bercic-mass}. However, one should be accurate with the definition of velocities because two
different phases co-exist in the bubble train flow. Usually, one can take the velocity $U$ to be
a bulk velocity or $U=\ugas+\uliq$, where $\ugas$ and $\uliq$ are liquid and gas
superficial velocities, respectively. {\color{red} The gas and liquid superficial and actual velocities are connected with each other through a complex dependency on $Ca$, $\epsilon$ \cite{kuzmin-binary3d}.}

With experimental measurements of concentration at different locations, the calculation
of the mass transfer coefficient using the logarithmic function is straightforward.
However, if one wants to analytically or numerically calculate the mass transfer coefficients, the
situation is much more complicated because of the presence of two phases and complex bubble
geometry. As was mentioned before, depending on the capillary
number the velocity pattern and thus scalar mixing is different. Analytical approaches
\cite{irandoust,vanbaten-circular} assume that the
contributions from film and bubble caps can be calculated separately. Therefore, tracer from the film does not influence bubble caps diffusion.   However, this assumption overpredicts mass transfer for a number of experiments \cite{irandoust}. This happens since some tracer
concentration from the film is mixed with the slug and increases the overall concentration in the slug, thereby decreasing
the mass transfer from the bubble caps.
Therefore, the analytical estimates for the mass transfer coefficient calculation  do not account for mutual mass
transfer from neighbouring bubbles.

@@

We perform simulations in the frame co-moving with the bubble
in which the bubble position stays constant. The bubble velocity $\ububble$ is
different from the bulk velocity $U=\ugas+\uliq$, and one thus needs to perform a $x$ coordinate
variable change:
\beqal
\label{theor:average:concentration:time}
&x(t)=\ububble t\\
&\overline{C(x)}=\cstar \Bigl(1-\exp\bigl(-\vol \frac{x}{\ugas+\uliq}\bigr)\Bigr)\\
&\langle C(t)\rangle=\cstar \Bigl(1-\exp\bigl(-\vol\, t \frac{\ububble}{\ugas+\uliq}\bigr)\Bigr),
\feqal
where $\langle C(t)\rangle$ is the space-averaged characteristic concentration,
and $\overline{C(x)}$ is the time-averaged concentration at location $x$. 
%The geometry is not one-dimensional and one needs the characteristic concentration
%to be only as a function of time. 
One can make different choices for
$\langle C(t) \rangle$ such as the concentration averaged over the whole domain or inlet/outlet space-averaged
concentrations as used in \cite{vanbaten-circular,kreutzer-overview}.  
%Note that $C(x)$ is not averaged in time as $\overline{C(x)}$ since $C(x)$
%ideally should not depend on time. 
The volumetric mass transfer coefficient can be obtained through the space-averaged concentration:
\beqal
\label{theor:one:concentration:time}
&\vol\, t \frac{\ububble}{\ugas+\uliq}=\ln \frac{\cstar}{\cstar-\langle C(t)\rangle}\\
&\volnondim=\frac{\lunit}{\ububble t}\ln \frac{\cstar}{\cstar-\langle C(t) \rangle},
\feqal
where the parameter $\vol \frac{\lunit}{\ugas+\uliq}$ is non-dimensional. One can also measure the
volumetric mass transfer coefficient from concentrations given at times $t_1$ and $t_2$:
\beq
\label{theor:continuous:mass:transfer}
\volnondim=\frac{\lunit}{\ububble
(t_2-t_1)}\ln\frac{C^{*}-\langle C(t_1) \rangle}{C^{*}-\langle C(t_2) \rangle}.
\feq
Expressions (\ref{theor:average:concentration:time} - \ref{theor:continuous:mass:transfer}) are the
cornerstones of the present work . Four possible scenarios of numerical simulations have been examined: 
\begin{enumerate}
\item % The bubble train multiphase simulations are conducted by applying periodic boundary
%conditions \cite{kuzmin-binary2d,kuzmin-binary3d}. 
%One
%can make an assumption of same boundary
%conditions applied to the tracer concentration to simulate mass transfer. 
One unit
cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. In this case no tracer leaves the
domain similarly to plug flow. 
%From the physical point of view this case resembles the mass transfer for small Capillary numbers
%$Ca<0.1$.  The liquid slug is well mixed and one have a homogeneous concentration in the unit
%cell, i.e. the outlet and the inlet concentrations are close. 
Though easier to implement, it gives rise to the criticism
that the inlet concentration is equal to the outlet one. As was
discussed, in experiments there is a concentration difference between the
inlet and the outlet, even for one unit cell.

In this case, the volumetric mass transfer coefficient is calculated by Eq. \ref{theor:one:concentration:time}.
The
characteristic concentration $\langle C(t) \rangle$ required for the volumetric mass
transfer coefficient is taken as the average concentration in the domain:
\beq
\label{average:concentration:definition}
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\feq
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{benchmark_middle.eps}\\
\caption{The two-dimensional benchmark for the  the mass transfer
coefficient for the bubble located at the middle of the domain. \label{fig:benchmark}}
\end{figure}
\item Periodic boundary conditions are applied as in the first case but the characteristic concentration is taken as the inlet/outlet flux-averaged
concentration \cite{vanbaten-circular}:
\beqal
&\langle \cinlet(t) \rangle=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
&\langle \coutlet(t) \rangle=\frac{\int{U(y) C(\lunit,y,t) \mathrm{d}y}}{\int{U(\lunit,y)
\mathrm{d}y}}\\
&\cinlet(\bm{x},t)=\coutlet(\bm{x},t),\,\text{due to periodicity}.
\feqal
The assumptions of this approach are
that the concentration difference between the inlet/outlet- and the space-averaged over the whole unit cell is not significant. Thus, the tracer is assumed to be well mixed in the slug.
\item
The approach of \citet{vanbaten-circular}, where periodic boundary conditions are used and
the mass transfer coefficient is calculated as the gain of the mass in the system
divided by the concentration difference
multiplied by the surface area:
\beq
\label{eq:vanbaten:formulation}
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-\langle C(t) \rangle)},
\feq
where the mass flux in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(\bm{x},t_2)\mathrm{d}\bm{x}}-\int_{liq}{
C(\bm{x},t_1)\mathrm { d } \bm{x}} } {
t_2-t_1 }.
\feq
In the approach of \citeauthor{vanbaten-circular} the inlet/outlet flux-averaged concentrations were taken as the
characteristic concentration $\langle C(t) \rangle $.

@@

With periodic boundary conditions: 
\beqstar
\int{\coutlet(t)
u(\lunit,y)\mathrm{d}y}=\int{\cinlet(t) u(0,y)\mathrm{d}y}. 
\feqstar
In this case, Eq. \ref{main:main:main} reduces to Eq. \ref{eq:vanbaten:formulation}.

%Then the mass change in the domain arises only because of the bubble
%surface. However, to calculate the mass transfer coefficient one needs to have
%the characteristic concentration  $\langle C(t) \rangle$ and the mass gain $\dot{m}$
%which do require the storage of both temporal and spatial information.
Another limiting case (will be shown later) is when the mass accumulation rate equals zero, i.e. $\dot{m}=0$. 
This situation corresponds to a simulation of a few unit cells with open boundary for flows with
$Ca>0.7$.   
%One
%can use only spatial information to calculate the mass transfer coefficient,
%i.e. $\int{\cinlet(t) u(0,y) \mathrm{d}y}$ and $\int{\coutlet(t)
%u(\lunit,y)\mathrm{d}y}$ require only space information.

@@

\subsection{TRT $D2Q9$ model}
The lattice Boltzmann equation (LBE) operates on a square/cubic grid representing the
physical domain {\color{red} in non-dimensional terms} (hereafter we use subscript LB to denote non-dimensional quantities). It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. The LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along  
directions specified by a discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)= f_i(\bm{x},t)-\omega \bigl(f_i(\bm{x},t)-eq_i(\bm{x},t)\bigr),&&\text{
collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $eq_i$ is the equilibrium probability distribution function, and $\omega$ is the
relaxation parameter. The term $-\omega \bigl(f_i- eq_i\bigr)$ is the so-called BGK
collision operator \cite{bgk}. However, the approach used here is the TRT
(two-relaxation-times) collision operator \cite{ginzburg-main,ginzburg-saturated-flow}. In
comparison with the widely used BGK collision operator, the TRT collision operator has better accuracy
for diffusion and convection fluxes, as well as a larger range of parameters where the scheme is stable.
